---
title: "Google transit data across the Stockholm region. Or: Show me where you live and I tell you if you work from home."
author: "Dominik Freunberger (https://dmnkfr.github.io/about.html)"
date: "3/9/2021"
output: html_document
---
A [friend](https://www.andifugard.info/) recently published [a very nice analysis](https://inductivestep.github.io/Google-transit-London/) of how Londoners changed their mobility behavior during the COVID-19 pandemic as compared to before. I wondered how this looks for the Stockholm region in Sweden.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r message = F}
library(tidyverse)
library(magrittr)
library(lubridate)
library(readxl)
library(ggrepel)
library(ggpubr)
```
The data is from [Google](https://www.google.com/covid19/mobility/), so, to quote [Andy](https://inductivestep.github.io/Google-transit-London/), "data quality unclear..."

``` {r include = T}
if(!file.exists("movement_data")) {
        dir.create("movement_data")
}
# dataURL = "https://www.gstatic.com/covid19/mobility/Region_Mobility_Report_CSVs.zip"
# download.file(dataURL, destfile="./movement_data/movement.zip", method = "curl")
# unzip ("./movement_data/movement.zip", exdir = "./movement_data")
```

Let's read the Swedish (SE) data, get the data for the Stockholm region, and tidy it up a bit:
``` {r include = T}
data = read.csv("./movement_data/2020_SE_Region_Mobility_Report.csv")

data %<>% mutate(
        day = parse_date(date, "%Y-%m-%d"),
        day_of_week = wday(day, label = TRUE),
        weekday = ifelse(day_of_week %in% c("Sat", "Sun"), "Weekend", "Weekday")
        )

sthlm_data <- data %>%
        filter(sub_region_1 == "Stockholm County" & sub_region_2 != "") 

sthlm_data %>%
        group_by(sub_region_2) %>%
        tally()

sthlm_data %>%
        select(ends_with("percent_change_from_baseline")) %>%
        names()

tidy_sthlm <- sthlm_data %>%
        pivot_longer(
          cols = ends_with("percent_change_from_baseline"),
          names_to = "location",
          names_pattern = "(.+)_percent_change_from_baseline",
          values_to = "percent_change_from_baseline")
        
tidy_sthlm$sub_region_2 = word(tidy_sthlm$sub_region_2, 1)

tidy_sthlm %>%
        select(location, percent_change_from_baseline) %>%
        filter(!is.na(percent_change_from_baseline)) %>%
        group_by(location) %>%
        summarise(mean_change = mean(percent_change_from_baseline, na.rm = T),
	          count = sum(!is.na(percent_change_from_baseline)),
	          NAs = sum(is.na(percent_change_from_baseline)))
```
As you can see, there's more data included but here we'll focus on the data about mobility trends for places of work. Let's grab that data and plot it:

```{r}
tidy_work = tidy_sthlm %>%
  filter(location %in% c("workplaces"))

plot1 <- tidy_work %>%
        mutate(location = gsub("_", " ", location)) %>%
        ggplot(aes(x = day, 
                   y = percent_change_from_baseline,
                   color = sub_region_2)) +
        theme(legend.position = "none") +
        labs(x = "Date", y = "% Change from Baseline") +
        theme_minimal()

```

```{r fig.height=4, fig.width=6, dpi = 300}
plot1 +
  geom_smooth(se = F, size = 1.2) +
        scale_color_discrete(name = "Municipality")
```

Which areas have more people who still have to travel to their workplace?

```{r fig.height=8, fig.width=6, dpi = 300}
tidy_sthlm %>%
        filter(location %in% c("workplaces")) %>%
        filter(day >= as.Date("2021-01-01") &
                       day < as.Date("2021-03-01")) %>%
        group_by(sub_region_2) %>%
        summarise(mean_change = mean(percent_change_from_baseline)) %>%
        na.omit() %>%
        mutate(sub_region_2 = fct_reorder(sub_region_2, mean_change)) %>%
        ggplot(aes(y = sub_region_2, x = mean_change)) +
        geom_point(size = 3, col = "#e02c24") +
        labs(y = "Municipality", 
             x = "% Change from baseline", 
             title = "Visits to workplaces",
             subtitle = "Mean change from pre-Covid baseline, Jan to Feb 2021") +
        theme_minimal()
```

## __Interesting.__   
You don't have to be living in Stockholm for long to see that this order is not random. Rather, it seems to have to do with how posh, or, how wealthy a municipality is. Let's get the Swedish income data from [SCB](https://www.scb.se/hitta-statistik/statistik-efter-amne/hushallens-ekonomi/inkomster-och-inkomstfordelning/inkomster-och-skatter/pong/tabell-och-diagram/inkomster--individer-lankommun/sammanraknad-forvarvsinkomst-2019--per-kommun-efter-percentiler/).

``` {r include = F}
if(!file.exists("income_data")) {
        dir.create("income_data")
}
# I downloaded the data here:
# https://www.scb.se/hitta-statistik/statistik-efter-amne/hushallens-ekonomi/inkomster-och-inkomstfordelning/inkomster-och-skatter/

```

Read in the data and grab the data for the Stockholm area:
``` {r include = T}
income = read_excel("./income_data/income.xls", skip = 3)
income = income[,c(1,2,4)]
income = income %>%
        rename(municipality = ...1, 
               mean_income = `Kvinnor och män`,
               median_income = ...4)

sthlm_income = income %>%
        filter(municipality %in% tidy_sthlm$sub_region_2)

sthlm_work = tidy_sthlm %>%
        filter(location %in% c("workplaces")) %>%
        filter(day >= as.Date("2021-01-01") & day < as.Date("2021-03-01")) %>%
        group_by(sub_region_2) %>%
        summarise(mean_change = mean(percent_change_from_baseline)) %>%
        na.omit()

sthlm_work$sub_region_2 = word(sthlm_work$sub_region_2, 1)

sthlm_work = sthlm_work %>%
        rename(municipality = sub_region_2) %>%
        inner_join(sthlm_income, by = "municipality")

sthlm_work$mean_income = round(as.numeric(sthlm_work$mean_income), 2)
sthlm_work$median_income = round(as.numeric(sthlm_work$median_income), 2)
```



```{r fig.height=6, fig.width=6, dpi = 300}
ggplot(sthlm_work, aes(x = mean_change, y = median_income)) +
    geom_point(size= 3) +    
  stat_smooth(method = "lm",
        col = "#e02c24",
        se = T,
        size = 2) +
        labs(title = "Workplace Mobility: A Matter of Wealth",
             x = "% Mobility Change from Baseline",
             y = "Median Income of Municipality in SEK (2019)") +
                ggrepel::geom_text_repel(aes(label = municipality), size = 3) +
        theme_minimal() +
        stat_cor(method="pearson", label.x = -28, label.y = 370000, size = 5)
```

So, one could say people in these areas care less about restrictions in general and travel more anyway. How then does mobility in other domains, such as grocery shopping or recreation look in relation to the municipalities' wealth?   

``` {r include = T}
sthlm_recr = tidy_sthlm %>%
        filter(location %in% c("retail_and_recreation")) %>%
        filter(day >= as.Date("2021-01-01") & day < as.Date("2021-03-01")) %>%
        group_by(sub_region_2) %>%
        summarise(mean_change = mean(percent_change_from_baseline)) %>%
        na.omit()

sthlm_recr$sub_region_2 = word(sthlm_recr$sub_region_2, 1)

sthlm_recr = sthlm_recr %>%
        rename(municipality = sub_region_2) %>%
        inner_join(sthlm_income, by = "municipality")

sthlm_recr$mean_income = round(as.numeric(sthlm_recr$mean_income), 2)
sthlm_recr$median_income = round(as.numeric(sthlm_recr$median_income), 2)
```

```{r fig.height=6, fig.width=6, dpi = 300}
ggplot(sthlm_recr, aes(x = mean_change, y = median_income)) +
    geom_point(size= 3) +
    stat_smooth(method = "lm",
        col = "#e02c24",
        se = T,
        size = 2) +
        labs(title = "Retail and recreation mobility as a function of income",
             x = "% Mobility Change from Baseline",
             y = "Median Income of Municipality in SEK (2019)") +
                ggrepel::geom_text_repel(aes(label = municipality), size = 3) +
        theme_minimal() +
        stat_cor(method="pearson", label.x = -10, label.y = 390000, size = 5)
```
``` {r include = T}
sthlm_groc = tidy_sthlm %>%
        filter(location %in% c("grocery_and_pharmacy")) %>%
        filter(day >= as.Date("2021-01-01") & day < as.Date("2021-03-01")) %>%
        group_by(sub_region_2) %>%
        summarise(mean_change = mean(percent_change_from_baseline)) %>%
        na.omit()

sthlm_groc$sub_region_2 = word(sthlm_groc$sub_region_2, 1)

sthlm_groc = sthlm_groc %>%
        rename(municipality = sub_region_2) %>%
        inner_join(sthlm_income, by = "municipality")

sthlm_groc$mean_income = round(as.numeric(sthlm_groc$mean_income), 2)
sthlm_groc$median_income = round(as.numeric(sthlm_groc$median_income), 2)
```

```{r fig.height=6, fig.width=6, dpi = 300}
ggplot(sthlm_groc, aes(x = mean_change, y = median_income)) +
    geom_point(size= 3) +
    stat_smooth(method = "lm",
        col = "#e02c24",
        se = T,
        size = 2) +
        labs(title = "Grocery and pharmacy mobility as a function of income",
             x = "% Mobility Change from Baseline",
             y = "Median Income of Municipality in SEK (2019)") +
                ggrepel::geom_text_repel(aes(label = municipality), size = 3) +
        theme_minimal() +
        stat_cor(method="pearson", label.x = -10, label.y = 390000, size = 5)
```


